# Tidy your data

## what is tidy data?

Three rules:

* Each variable must have its own column.

* Each observation must have its own row.

* Each value must have its own cell.

```{r message=FALSE}
library(tidyverse)
library(knitr)
# tidy data
kable(table1, caption = "a This is a tidy dataset!")

kable(table2, caption = "b case and population missing")

kable(table3, caption = "c case and population merged")

kable(table4a, caption = "d year missing, cases")

kable(table4b, caption = "e year missing, population")
```

## Tools in tidyr 

### gather ( )

For some datasets, some of the column names are not names of variables but **values** of a variable.

```{r}
kable(table4a, caption= "year missing, cases")

```


We need to gather years like 1999 and 2000 into one variable **year**.

```{r}

gather(table4a, '1999','2000',key = "year", value = "cases")

gather(table4b, '1999','2000',key = "year", value = "population")

```


### spread ( )
Sometime the observations are scattered across muiltiple rows.

```{r}

kable(table2, caption = "b case and population missing")

spread(table2, key = type, value = count)

```

To sum up, gather() makes wide tables long while spread() makes long tables wide. The key idea here is to think about the variables you want to explore and make sure that they are placed in different columns. This is very important for data visualization and stats modeling in R.

***


### separate()

In some cases, one column may contain a combination of two variables. We need to separate these two variables.

```{r}
kable(table3, caption = "c case and population merged")

#separate by a delimiter
separate(table3, rate, into = c("cases", "population"), sep = "/")

#separate by the first two digits

separate(table3, rate, into = c("cases", "population"), sep = 2)

#separate by the last two digits

separate(table3, rate, into = c("cases", "population"), sep = -2)

```

### unite()

```{r}
table_new = separate(table3, year, into = c("year1", "year2"), sep = 2)

table_new

unite(table_new, new, year1, year2)# by default there is an underscore between two parts.

unite(table_new, new, year1, year2, sep = "")
```
