# Tranform your data

## Selecting variables of interest
When we get a dataset, one of the most important things to do is to pick the variables we need, especially when the raw dataset has too many columns.

```{r message=FALSE}
library(tidyverse)

# getting the data
filename = list.files(paste0(getwd(), "/data"),full.names=T, pattern = ".csv") # get only csv files

df = data.frame()
bin = data.frame()
for (i in 1:length(filename)){
  # length(filename) = how many files
  bin = read.csv(filename[i])
  df = rbind(df, bin)
}

#head(df)

#colnames(df)

```

In this dataset *df*, we need to get the *ID* of participants as stored in Subject, the *stimuli* they hear under the column name "tone.Trial", their *response* "insex1.RESP", their *response time* in "insex1.RT", and the *experiment conditions* in "Procedure.Block.".

```{r}

df_new = select(df, subject = "Subject",
                stimuli = "tone.Trial.",
                response = "insex1.RESP",
                response_rt = "insex1.RT",
                block = "Procedure.Block.",
                ExperimentName)
head(df_new)

# select by names
head(select(df_new, subject: response))
head(select(df_new, subject: response))

head(select(df_new, 1: 3))

head(select(df_new, -subject))

head(select(df_new, ExperimentName, everything()))

# change column names

head(rename(df_new, exp = ExperimentName))

```




## Filtering observations

Not all the data in the raw dataset are useful. We can use filter() function to select rows or observation based on some criteria.

Three useful logical operators for forming conditions:

 + **&** means "and"
 
 + **|** means "or"
 
 + **!** means "not"

```{r}
# removing data from the practice block
df_new = filter(df_new, block != "pracproc")
head(df_new)
# removing missing data in response column
df_new = filter(df_new, !is.na(response))
df_new = filter(df_new, response != "")
head(df_new)

# pick up stimuli 33 and 21

head(filter(df_new, stimuli == 33 | stimuli == 21))

head(filter(df_new, stimuli %in% c(33, 21)))

# removing data with respones time shorter than 200

head(filter(df_new, response_rt < 200))

head(filter(df_new, response_rt > 200))

```

## changing the order of rows

```{r}
# arrange the dataframe by stimuli and response_rt columns
head(arrange(df_new, stimuli, response_rt))

# arrange the dataframe by stimuli and the descending order of block columns
head(arrange(df_new, stimuli, desc(block)))



```

## Generating new variables

We can generate new variable based on existing variables using the function mutate() and transmutate(). The difference is mutate will keep the original variables while transmutate will erase orginal variables.

```{r}
# you can add, minus, log transform any numberic column
head(mutate(df_new, response_rt_new = log(response_rt)))

# you can extract part of the information from a character variable

head(mutate(df_new, ISI = str_extract(ExperimentName, "2000|500")))

head(transmute(df_new, ISI = str_extract(ExperimentName, "2000|500")))

# you can paste two variabls together

head(transmute(df_new, ISI = paste(stimuli, block, sep = "_")))

# you can recode a variable

head( mutate(df_new, block = recode(block, 
                         block1 = "ss", block2 = "ss",
                         block3 = "sd", block4 = "sd",
                         block5 = "ds", block6 = "ds",
                         block7 = "dd")))

```
## the pipeline “%>%”
You may have noticed the repetition in the above codes. We have to specify the dataframe every time we use a function and have to store it in a new name if we want to keep the original dataframe.

Using pipeline *%>%* can reduce the redundancy and the mistake we may make when we have a dozen of dataframes.


```{r}
df_final = df %>%
  select(., subject = "Subject",
                stimuli = "tone.Trial.",
                response = "insex1.RESP",
                response_rt = "insex1.RT",
                block = "Procedure.Block.",
                exp = ExperimentName)%>%
  filter(block != "pracproc" & !is.na(response) & response != "")%>%
  mutate(ISI = str_extract(exp, "2000|500"),
         block = recode(block, 
                         block1 = "ss", block2 = "ss",
                         block3 = "sd", block4 = "sd",
                         block5 = "ds", block6 = "ds",
                         block7 = "dd"))
           
  
  

```

## Summarizing 

Summarize() and group_by () are often used together to give us some basic summary of the data.

```{r}
df_final%>%
  group_by(subject)%>%
  summarize(count = n())

df_final%>%
  group_by(subject, stimuli)%>%
  summarize(rt = mean(response_rt))

df_final%>%
  group_by(stimuli, response)%>%
  mutate(counter = 1)%>%
  summarize(count = sum(counter))%>%
  spread(stimuli, value = count)


df_final%>%
  group_by(stimuli,response)%>%
  mutate(counter = 1)%>%
  summarize(counter = sum(counter))%>%
  mutate( percentage = counter/sum(counter),
          sum = sum(counter))
 

table.final = df_final%>%
  group_by(stimuli,response)%>%
  mutate(counter = 1)%>%
  summarize(counter = sum(counter))%>%
  mutate( percentage = counter/sum(counter),
          sum = sum(counter))%>%
  select(stimuli,percentage, response)%>%
  spread(stimuli, value = round(percentage, 2))

library(knitr)
kable(table.final, caption = "Percentage of choice")

```
